$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-12-15 paper/paper.tex
A much more to-the-point introduction

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{
  $\pi$ with leftovers: \\ a mechanisation in Agda
  \thanks{The authors would also like to thank the anonymous referees for their valuable comments and helpful suggestions. This work is supported by the UK EPSRC grant EP/K034413/1 ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233 ``Behavioural Application Program Interfaces'' (BehAPI).}
}
\titlerunning{$\pi$ with leftovers: a mechanisation in Agda}
\author{
  Uma Zalakain\inst{1}\orcidID{0000-0002-3268-9338}
  \and
  Ornela Dardha\inst{2}\orcidID{0000-0001-9927-7875}
}
\institute{
  University of Glasgow, Scotland \\ \email{u.zalakain.1@research.gla.ac.uk}
  \and
  University of Glasgow, Scotland \\ \email{ornela.dardha@glasgow.ac.uk}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
  \todo{the only references here are to session types}
  \todo{fix acknowledgments}
  The \picalc{} is a computational model for communication and concurrency.
  The \emph{linear} \picalc{} is a typed version of the \picalc{} where channels must be used exactly once.
  It is an underlying theoretical and practical framework on top of which more advanced types and theories can be built, such as \emph{session types} \cite{H93,THK94,HVK98}.

  We present an untyped syntax and semantics for the \picalc{}, on top of which we provide a resource-aware type system parametrised by a set of multiplicity algebras.
  This allows the user to mix \emph{linear}, \emph{graded} and \emph{shared} types, under the same unified framework.
  We use leftover typing \cite{Allais2018a} to encode our typing rules, thus avoiding the ubiquitous context splitting proofs common when dealing with linear type systems.
  We provide a framing theorem, generalise the weakening and strengthening theorems, and use them to prove subject congruence.
  We show that the type system is stable under substitution and prove subject reduction.
%
  Our formalisation is fully mechanised in Agda.

  \keywords{Pi-calculus \and Linear types \and Leftover typing \and Concurrency \and Mechanisation \and Agda}
\end{abstract}

@@

The \picalc{} \cite{MilnerPW92,Milner99} is a computational model for communication and concurrency that boils concurrent processing down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first class values, and can therefore be sent and received.
The \picalc{} is at the basis of the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and more recently Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, session types, and types for liveness properties, such as deadlock freedom, livelock freedom or termination.
These make of the \picalc{} a fully-fledged tool for modelling and verifying concurrent and distributed systems.

In the early '90s, \emph{linear types} for (functional) programming languages \cite{Wadler90,Bernardy2018} were introduced, following J.-Y. Girard's development of linear logic \cite{Girard87}.
Linear type systems guarantee that resources are used \emph{exactly once}.
Enforcing this ``no duplicating'' and ``no discarding'' of resources via linearity allows for resource-aware programming and more efficient implementations \cite{Wadler90}.
Later on, linearity also inspired unique types, as in Clean \cite{BarendsenS96}, and ownership types, as in Rust \cite{MatsakisK14}.

\todo{introduce shared and graded types sooner}

Linearity influenced the $\pi$-calculus as well.
Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
With the rise of session types \cite{H93,THK94,HVK98} linearity acquired a different flavour: linearity ensures that each channel endpoint is owned by exactly one communicating participant, even though the channel itself may be used multiple times, as specified by its session type.
Linearity in session types is the key ingredient that guarantees communication safety and session fidelity.

In recent years, an encoding of session types into linear types \cite{DardhaGS12,Dardha14,DardhaGS17} has pushed the linear \picalc{} \cite{KPT96} into the spotlight again.
The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}, thus allowing to use the target linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be defined and implemented.
\todo{go back to linear \picalc{}}

In this paper we make the following contributions:
\begin{enumerate}
  \item We use \textbf{type-level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax for \picalc{} processes that is well-scoped by construction}:
    Every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
    \todo{Do Ciccone \& Padovani also use type level de Bruijn indices?}

  \item We \textbf{parametrise our type system over a set of resource-aware usage algebras}:
    A \emph{usage algebra} is defined as a ternary relation that is \emph{partial}, \emph{deterministic}, \emph{cancellative}, \emph{associative} and \emph{commutative}.
    The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
    This allows for linear, graded and shared types to be mixed in the same type system.

  \item We \textbf{apply leftover typing to the \picalc{} and prove type preservation for the resulting type system}:
    \todo{along the lines of Allais \cite{Allais2018a}}.
\end{enumerate}

\subsection{Summary}
\todo{shorten}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: we use type level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
   For convenience and readability, the user can convert from and to a syntax using channel names (\autoref{from_to_deBruijn}).
   These conversion functions are shown to be inverses of one another up to $\alpha$-conversion.

   \item \textbf{Semantics}: we provide an operational semantics for the \picalc{}, prior to any typing (\autoref{semantics}).
   Thanks to our well-scoped grammar  (\autoref{syntax}), the operational semantics is defined on \emph{the totality of the syntax} as a reduction relation on processes (\autoref{operational-semantics}).
   The reduction relation tracks at the type level the channel on which communication occurs. This information is later used to state the subject reduction theorem --- aka type preservation (\autoref{thm:subject-reduction}).
   The reduction relation is defined modulo \emph{structural congruence} (\autoref{structural-congruence}) --- a relation defined on processes that acts as a quotient type to remove unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
 \end{itemize}

  \item \textbf{Leftover typing for the \picalc{} with support for linear, graded and shared types}:
  we present our type system for a resource-aware \picalc{} in \autoref{type-system}.
  \begin{itemize}
    \item \textbf{Algebras on multiplicities}: The user can provide \emph{resource-aware} algebras, which are then applied to the type system (\autoref{multiplicities}).
    Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra --- linear types, graded types and shared types all satisfy these algebraic laws.

    Multiple algebras can be simultaneously used in a single type system --- usage contexts keep information about what algebra to use on which element (\autoref{contexts}).
    This allows for type systems combining linear, graded and shared types under the same framework.

    \item \textbf{Leftover typing}: Our type system uses \emph{leftover typing} to model the resource-aware \picalc{} (\autoref{leftover-typing}).
    This approach adds a leftover usage context to the typing judgements.
    Typing derivations take the resources of their input usage context, consume some of them, and leave the remaining as leftovers in the output usage context.
    Leftover typing \textbf{makes top-down context splits unnecessary}, allows for \emph{framing} (\autoref{thm:framing}) to be stated, and makes it possible to generalise \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) for the \picalc{}.
  \end{itemize}

  \item \textbf{Fully mechanised formalisation and type safety results}:
    The formalisation of the \picalc{} with leftover typing, from the syntax to the semantics and the type system, has been fully mechanised in Agda and is publicly available in our GitHub repository \cite{Zalakain2020Agda}

    We present the type safety results of our \picalc{} with leftovers in \autoref{type-safety}.
    We have fully mechanised \emph{framing} (\autoref{thm:framing}), \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}).
    We use these results to then prove and mechanise \emph{subject congruence} (\autoref{thm:subject-congruence}), and together with \emph{substitution} (\autoref{thm:substitution}), to prove and mechanise \emph{subject reduction} (\autoref{thm:subject-reduction}).
\end{enumerate}

@@

\section*{Acknowledgments}

We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{
  $\pi$ with leftovers: \\ a mechanisation in Agda
  \thanks{
    The authors would also like to thank the anonymous referees for their valuable comments and helpful suggestions.
    We further want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their work and thoughts.
    This work is supported by the UK EPSRC grant EP/K034413/1 ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233 ``Behavioural Application Program Interfaces'' (BehAPI).}
}
\titlerunning{$\pi$ with leftovers: a mechanisation in Agda}
\author{
  Uma Zalakain\inst{1}\orcidID{0000-0002-3268-9338}
  \and
  Ornela Dardha\inst{2}\orcidID{0000-0001-9927-7875}
}
\institute{
  University of Glasgow, Scotland \\ \email{u.zalakain.1@research.gla.ac.uk}
  \and
  University of Glasgow, Scotland \\ \email{ornela.dardha@glasgow.ac.uk}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
%  \todo{the only references here are to session types}
%  The \picalc{} is a computational model for communication and concurrency.
%  The \emph{linear} \picalc{} is a typed version of the \picalc{} where channels must be used exactly once.
%  It is an underlying theoretical and practical framework on top of which more advanced types and theories can be built, such as \emph{session types} \cite{H93,THK94,HVK98}.
%
%  We present an untyped syntax and semantics for the \picalc{}, on top of which we provide a resource-aware type system parametrised by a set of multiplicity algebras.
%  This allows the user to mix \emph{linear}, \emph{graded} and \emph{shared} types, under the same unified framework.
%  We use leftover typing \cite{Allais2018a} to encode our typing rules, thus avoiding the ubiquitous context splitting proofs common when dealing with linear type systems.
%  We provide a framing theorem, generalise the weakening and strengthening theorems, and use them to prove subject congruence.
%  We show that the type system is stable under substitution and prove subject reduction.
%%
%  Our formalisation is fully mechanised in Agda.
%
  \keywords{Pi-calculus \and Linear types \and Leftover typing \and Concurrency \and Mechanisation \and Agda}
\end{abstract}

@@

Recent years have seen an increase in the efforts to mechanise resource-aware process algebras \todo{cite}.
A resource-aware type system needs to keep track of what resources are used in which parts of the program.
To do so, the standard approach is to use context splits\todo{cite}: a typing rule for a term with multiple subterms will require an extra arguments that specifies what resources to allocate to each of the subterms.
The typing derivations for the subterms must then use the entirety of resources allocated to them.
A key observation here is that each subterm already \emph{knows} about the resources it needs, especially if these are linear.
\emph{Context splits do little more than repeat information that is already contained in the subterms.}
Moreover, it is impossible to provide typing derivations for the subterms until after the context splits have been provided.
On top of that, because in their usual presentation context splits are binary, a typing rule for a term with $n$ subterms requires $n - 1$ context splits, which adds to the clutter of the type system.

An alternative approach is to using context splits is using leftover typing \todo{cite}.
Leftover typing changes the form of the typing judgments themselves by including a second output \emph{leftover} context that contains the resources that were unused by the term.
As a result, typing rules can now \emph{thread the resources through} subterms without the need of context splits.
Each subterm uses the resources it needs, and leaves the rest for its siblings.
The first subterm in this chain of resources immediately knows about the resources it has available.

Leftover typing has been used to formulate intuitionistic linear logic \todo{cite} and to mechanise the linear \lambdacalc.
We make two main contributions in this paper:
\begin{itemize}
  \item We use \textbf{leftover typing to type a resource-aware \picalc{} and prove subject reduction}.
  \item We \textbf{generalise the resource counting to a set of usage algebras that can be mixed within the same type system}.
\end{itemize}

% The \picalc{} \cite{MilnerPW92,Milner99} is a computational model for communication and concurrency that boils concurrent processing down to the sending and receiving of data over communication channels.
% Notably, it features channel mobility: channels themselves are first class values, and can therefore be sent and received.
% The \picalc{} is at the basis of the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and more recently Go \footnote{https://golang.org}.
% At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, session types, and types for liveness properties, such as deadlock freedom, livelock freedom or termination.
% These make of the \picalc{} a fully-fledged tool for modelling and verifying concurrent and distributed systems.
% 
% In the early '90s, \emph{linear types} for (functional) programming languages \cite{Wadler90,Bernardy2018} were introduced, following J.-Y. Girard's development of linear logic \cite{Girard87}.
% Linear type systems guarantee that resources are used \emph{exactly once}.
% Enforcing this ``no duplicating'' and ``no discarding'' of resources via linearity allows for resource-aware programming and more efficient implementations \cite{Wadler90}.
% Later on, linearity also inspired unique types, as in Clean \cite{BarendsenS96}, and ownership types, as in Rust \cite{MatsakisK14}.
% 
% \todo{introduce shared and graded types sooner}
% 
% Linearity influenced the $\pi$-calculus as well.
% Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
% %
% With the rise of session types \cite{H93,THK94,HVK98} linearity acquired a different flavour: linearity ensures that each channel endpoint is owned by exactly one communicating participant, even though the channel itself may be used multiple times, as specified by its session type.
% Linearity in session types is the key ingredient that guarantees communication safety and session fidelity.
% 
% In recent years, an encoding of session types into linear types \cite{DardhaGS12,Dardha14,DardhaGS17} has pushed the linear \picalc{} \cite{KPT96} into the spotlight again.
% The encoding has been used as a technique to implement session types in mainstream programming languages such as OCaml \cite{Padovani17} and Scala \cite{ScalasY16,ScalasDHY17}, thus allowing to use the target linear \picalc{} as an underlying theoretical and practical framework on top of which session types can be defined and implemented.
% \todo{go back to linear \picalc{}}
% 
% In this paper we make the following contributions:
% \begin{enumerate}
%   \item We use \textbf{type-level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax for \picalc{} processes that is well-scoped by construction}:
%     Every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
%     \todo{Do Ciccone \& Padovani also use type level de Bruijn indices?}
% 
%   \item We \textbf{parametrise our type system over a set of resource-aware usage algebras}:
%     A \emph{usage algebra} is defined as a ternary relation that is \emph{partial}, \emph{deterministic}, \emph{cancellative}, \emph{associative} and \emph{commutative}.
%     The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
%     This allows for linear, graded and shared types to be mixed in the same type system.
% 
%   \item We \textbf{apply leftover typing to the \picalc{} and prove type preservation for the resulting type system}:
%     \todo{along the lines of Allais \cite{Allais2018a}}.
% \end{enumerate}
% 
% 
% 
% \subsection{Summary}
% \todo{shorten}
% \begin{enumerate}
% \item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
%  \begin{itemize}
%    \item \textbf{Syntax}: we use type level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
%    For convenience and readability, the user can convert from and to a syntax using channel names (\autoref{from_to_deBruijn}).
%    These conversion functions are shown to be inverses of one another up to $\alpha$-conversion.
%    
%    \item \textbf{Semantics}: we provide an operational semantics for the \picalc{}, prior to any typing (\autoref{semantics}).
%    Thanks to our well-scoped grammar  (\autoref{syntax}), the operational semantics is defined on \emph{the totality of the syntax} as a reduction relation on processes (\autoref{operational-semantics}).
%    The reduction relation tracks at the type level the channel on which communication occurs. This information is later used to state the subject reduction theorem --- aka type preservation (\autoref{thm:subject-reduction}).
%    The reduction relation is defined modulo \emph{structural congruence} (\autoref{structural-congruence}) --- a relation defined on processes that acts as a quotient type to remove unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
%  \end{itemize}
%   
%   \item \textbf{Leftover typing for the \picalc{} with support for linear, graded and shared types}:
%   we present our type system for a resource-aware \picalc{} in \autoref{type-system}.
%   \begin{itemize}
%     \item \textbf{Algebras on multiplicities}: The user can provide \emph{resource-aware} algebras, which are then applied to the type system (\autoref{multiplicities}).
%     Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra --- linear types, graded types and shared types all satisfy these algebraic laws.
% 
%     Multiple algebras can be simultaneously used in a single type system --- usage contexts keep information about what algebra to use on which element (\autoref{contexts}).
%     This allows for type systems combining linear, graded and shared types under the same framework.
%     
%     \item \textbf{Leftover typing}: Our type system uses \emph{leftover typing} to model the resource-aware \picalc{} (\autoref{leftover-typing}).
%     This approach adds a leftover usage context to the typing judgements.
%     Typing derivations take the resources of their input usage context, consume some of them, and leave the remaining as leftovers in the output usage context.
%     Leftover typing \textbf{makes top-down context splits unnecessary}, allows for \emph{framing} (\autoref{thm:framing}) to be stated, and makes it possible to generalise \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) for the \picalc{}.
%   \end{itemize}
% 
%   \item \textbf{Fully mechanised formalisation and type safety results}:
%     The formalisation of the \picalc{} with leftover typing, from the syntax to the semantics and the type system, has been fully mechanised in Agda and is publicly available in our GitHub repository \cite{Zalakain2020Agda}
% 
%     We present the type safety results of our \picalc{} with leftovers in \autoref{type-safety}.
%     We have fully mechanised \emph{framing} (\autoref{thm:framing}), \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}).
%     We use these results to then prove and mechanise \emph{subject congruence} (\autoref{thm:subject-congruence}), and together with \emph{substitution} (\autoref{thm:substitution}), to prove and mechanise \emph{subject reduction} (\autoref{thm:subject-reduction}).
% \end{enumerate}

@@
